\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-02-06 analysis/paper/paper.tex
comparison of casebase to survreg for case study 1

\usepackage{amsmath} \usepackage{longtable} \usepackage{graphicx} \usepackage{tabularx} \usepackage{float} \usepackage{booktabs} \usepackage{makecell} \DeclareUnicodeCharacter{2500}{-}

@@

@@

\begin{CodeOutput}
#>           (Intercept) ScrArmScreening group 
#>                 -7.77                 -0.22
\end{CodeOutput}

\begin{CodeOutput}
#>           (Intercept) ScrArmScreening group 
#>                 -7.76                 -0.23
\end{CodeOutput}

\begin{CodeOutput}
#>           (Intercept) ScrArmScreening group 
#>                -10.17                 -0.22
\end{CodeOutput}

\begin{CodeOutput}
#>           (Intercept)   log(Follow.Up.Time) ScrArmScreening group 
#>                 -9.44                  1.07                 -0.23
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{table}[H]
\centering
\begin{tabular}{lrlr}
\toprule
Model & HR & CI & Width\\
\midrule
Cox & 0.80 & (0.67, 0.95) & 0.28\\
Exponential & 0.81 & (0.68, 0.96) & 0.28\\
Gompertz & 0.78 & (0.66, 0.93) & 0.27\\
Weibull & 0.80 & (0.67, 0.95) & 0.28\\
Splines & 0.81 & (0.68, 0.96) & 0.28\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{CodeOutput}
#> Analysis of Deviance Table
#> 
#> Model 1: DeadOfPrCa ~ ScrArm + offset(offset)
#> Model 2: DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm + offset(offset)
#>   Resid. Df Resid. Dev Df Deviance            Pr(>Chi)    
#> 1     54538       6053                                    
#> 2     54535       5786  3      267 <0.0000000000000002 ***
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{CodeOutput}
\end{CodeChunk}

As expected, we see that splines model provides a better fit. Similarly,
we can compare the AIC for all three case-base models:

\begin{CodeOutput}
#>    Exp.   Gomp.   Weib. Splines 
#>    6057    5816    6057    5796
\end{CodeOutput}
\end{CodeChunk}

Once again, we have evidence that the splines model provides the best
fit.

With this first case study, we saw how population-time plots can provide
a useful summary of the incidence for our event of interests. In
particular, we saw how it can provide evidence against an exponential
hazard model. We then explored how different parametric models can
provide different estimates of the CIF, even though they all give
similar estimates for the hazard ratios.

\hypertarget{case-study-2bone-marrow-transplant}{%
\section{Case study 2--Bone-marrow
transplant}\label{case-study-2bone-marrow-transplant}}

In the next case study, we show how case-base sampling can also be used
in the context of a competing risk analysis. For illustrative purposes,
we will use the same data that was used in Scrucca \emph{et al}
\citeyearpar{scrucca2010regression}. The data was downloaded from the
first author's website, and it is also available as part of the
\pkg{casebase} package.

The data contains information on 177 patients who received a stem-cell
transplant for acute leukemia. The event of interest is relapse, but
other competing causes (e.g.~transplant-related death) were also
recorded. Several covariates were captured at baseline: sex, disease
type (acute lymphoblastic or myeloblastic leukemia, abbreviated as ALL
and AML, respectively), disease phase at transplant (Relapse, CR1, CR2,
CR3), source of stem cells (bone marrow and peripheral blood, coded as
BM+PB, or only peripheral blood, coded as PB), and age. A summary of
these baseline characteristics appear in Table \ref{tab:table1bmtcrr}.
We note that the statistical summaries were generated differently for
different variable types: for continuous variables, we gave the range,
followed by the mean and standard deviation; for categorical variables,
we gave the counts for each category.

\caption{\label{tab:table1bmtcrr}Baseline characteristics of patients in the stem-cell transplant study.}
\centering
\begin{tabular}[t]{lll}
\toprule
Variable & Description & Statistical summary\\
\midrule
\code{Sex} & Sex & M=Male (100)\\
 &  & F=Female (77)\\
\code{D} & Disease & ALL (73)\\
 &  & AML (104)\\
\code{Phase} & Phase & CR1 (47)\\
\addlinespace
 &  & CR2 (45)\\
 &  & CR3 (12)\\
 &  & Relapse (73)\\
\code{Source} & Type of transplant & BM+PB (21)\\
 &  & PB (156)\\
\addlinespace
\code{Age} & Age of patient (years) & 4--62\\
 &  & 30.47 (13.04)\\
\code{ftime} & Failure time (months) & 0.13--131.77\\
 &  & 20.28 (30.78)\\
\code{Status} & Status indicator & 0=censored (46)\\
\addlinespace
 &  & 1=relapse (56)\\
 &  & 2=competing event (75)\\
\bottomrule
\end{tabular}
\end{table}

First, we can look at a population-time plot to visualize the incidence
density of both relapse and the competing events. In Figure
\ref{fig:compPop}, failure times are highlighted on the plot using red
points for the event of interest (panel A) and blue dots for competing
events (panel B). In both panels, we see evidence of a non-constant
hazard function: the density of points is larger at the beginning of
follow-up than at the end.

{\centering \includegraphics{../figures/compPop-1}

\caption[Population-time plot for the stem-cell transplant study]{Population-time plot for the stem-cell transplant study. A) The red points represent relapse events. B) The blue points represent the competing events.}\label{fig:compPop}
\end{figure}
\end{CodeChunk}

Our main objective is to compute the absolute risk of relapse for a
given set of covariates. First, we fit a smooth hazard to the data using
a linear term for time:

We will compare our hazard ratio estimates to that obtained from a Cox
regression. To do so, we need to treat the competing event as censoring.

\begin{CodeInput}
R> library(survival)
R> # Treat competing event as censoring
R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
R+                    data = bmtcrr)
\end{CodeInput}
\end{CodeChunk}

\caption{\label{tab:bmtcrr-cis}ADD CAPTION}
\centering
\begin{tabular}[t]{lrlrl}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Case-Base} & \multicolumn{2}{c}{Cox} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
Covariates & HR & 95\% CI & HR & 95\% CI\\
\midrule
Sex & 0.80 & (0.46, 1.39) & 0.68 & (0.39, 1.19)\\
Disease & 0.58 & (0.32, 1.03) & 0.52 & (0.29, 0.93)\\
Phase (CR2 vs. CR1) & 1.16 & (0.46, 2.89) & 1.21 & (0.48, 3.02)\\
Phase (CR3 vs. CR1) & 1.73 & (0.44, 6.72) & 1.67 & (0.43, 6.5)\\
Phase (Relapse vs. CR1) & 3.93 & (1.82, 8.49) & 4.55 & (2.09, 9.9)\\
\addlinespace
Source & 1.65 & (0.54, 5.03) & 1.46 & (0.47, 4.54)\\
Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
\bottomrule
\end{tabular}
\end{table}

From the fit object, we can extract both the hazard ratios and their
corresponding confidence intervals. These quantities appear in Table
\ref{tab:bmtcrr-cis}. As we can see, the only significant hazard ratio
identified by case-base sampling is the one associated with the phase of
the disease at transplant. More precisely, being in relapse at
transplant is associated with a hazard ratio of 3.89 when compared to
CR1.

Given the estimate of the hazard function obtained using case-base
sampling, we can compute the absolute risk curve for a fixed covariate
profile. We perform this computation for a 35 year old woman who
received a stem-cell transplant from peripheral blood at relapse. We
compared the absolute risk curve for such a woman with acute
lymphoblastic leukemia with that for a similar woman with acute
myeloblastic leukemia. We will estimate the curve from 0 to 60 months.

We will compare our estimates to that obtained from a corresponding
Fine-Gray model \citep{fine1999proportional}. The Fine-Gray model is a
semiparametric model for the cause-specific \emph{subdistribution},
i.e.~the function \(f_k(t)\) such that
\[CI_k(t) =1 - \exp\left( - \int_0^t f_k(u) \textrm{d}u \right),\] where
\(CI_k(t)\) is the cause-specific cumulative indicence. The Fine-Gray
model allows to directly assess the effect of a covariate on the
subdistribution, as opposed to the hazard. For the computation, we will
use the \pkg{timereg} package \citep{timereg}.

\begin{CodeInput}
R> library(timereg)
R> model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) + 
R+                           const(Phase) + const(Source) + const(Age), 
R+                       data = bmtcrr, cause = 1, model = "fg")
R> 
R> # Estimate absolute risk curve
R> risk_fg <- predict(model_fg, newdata, times = time_points)
\end{CodeInput}
\end{CodeChunk}

{\centering \includegraphics{../figures/bmtcrr-risk-1}

\caption{\label{fig:compAbsrisk}Absolute risk curve for a fixed covariate profile and the two disease groups. The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:bmtcrr-risk}
\end{figure}
\end{CodeChunk}

Figure \ref{fig:compAbsrisk} shows the absolute risk curves for both
case-base sampling and the Fine-Gray model. As we can see, the two
approaches agree quite well for acute myeloblastic leukemia; however,
there seems to be a difference of about 5\% between the two curves for
acute lymphoblastic leukemia. This difference does not appear to be
significant: the curve from case-base sampling is contained within a
95\% confidence band around the Fine-Gray absolute risk curve (figure
not shown).

\hypertarget{case-study-3support-data}{%
\section{Case study 3--SUPPORT Data}\label{case-study-3support-data}}

In the first two case studies, we described the basic functionalities of
the \pkg{casebase} package: creating population-time plots, fitting
parametric models for hazard functions, and estimating the corresponding
cumulative incidence curves. In the next two case studies, we explore
more advanced features.

For the third case study, we will look at regularized estimation for the
hazard function. This can be performed using a combination of case-base
sampling and regularized logistic regression. To demonstrate this
capability, we examine the Study to Understand Prognoses Preferences
Outcomes and Risks of Treatment (SUPPORT) dataset
\citep{knaus1995support}. The SUPPORT dataset tracks death for
individuals who are considered seriously ill within a hospital.
Imputation was conducted using predictive mean matching and using the
\pkg{mice} package in \proglang{R} \citep{mice}. Before imputation,
there was a total of 1000 individuals and 35 variables. After
imputation, we have 474 individuals.
\textcolor{red}{How come we lose individuals during imputation?} Of
these, 67.5\% experienced the event of interest. A description of each
variable can be found in Table \ref{tab:support1} and a breakdown of
each categorical variable in Table \ref{tab:support2}. The original data
was obtained from the website of the Department of Biostatistics at
Vanderbilt University. The imputed dataset is available as part of the
\pkg{casebase} package.
\begin{tabular}{ccccc}
\textbf{Name} & \textbf{Labels}                          & \textbf{Levels} & \textbf{Storage} & \textbf{NAs} \\
age           & Age                                      &                 & double           & 0            \\
death         & Death at any time up to NDI date:31DEC94 &                 & double           & 0            \\
sex           &                                          & 2               & integer          & 0            \\
hospdead      & Death in Hospital                        &                 & double           & 0            \\
slos          & Days from Study Entry to Discharge       &                 & double           & 0            \\
d.time        & Days of Follow-Up                        &                 & double           & 0            \\
dzgroup       &                                          & 8               & integer          & 0            \\
dzclass       &                                          & 4               & integer          & 0            \\
num.co        & number of comorbidities                  &                 & double           & 0            \\
edu           & Years of Education                       &                 & double           & 202          \\
income        &                                          & 4               & integer          & 349          \\
scoma         & SUPPORT Coma Score based on Glasgow D3   &                 & double           & 0            \\
charges       & Hospital Charges                         &                 & double           & 25           \\
totcst        & Total RCC cost                           &                 & double           & 105          \\
totmcst       & Total micro-cost                         &                 & double           & 372          \\
avtisst       & Average TISS, Days 3-25                  &                 & double           & 6            \\
race          &                                          & 5               & integer          & 5            \\
meanbp        & Mean Arterial Blood Pressure Day 3       &                 & double           & 0            \\
wblc          & White Blood Cell Count Day 3             &                 & double           & 24           \\
hrt           & Heart Rate Day 3                         &                 & double           & 0            \\
resp          & Respiration Rate Day 3                   &                 & double           & 0            \\
temp          & Temperature (Celsius) Day 3              &                 & double           & 0            \\
pafi          & PaO2/(.01*FiO2) Day 3                    &                 & double           & 253          \\
alb           & Serum Albumin Day 3                      &                 & double           & 378          \\
bili          & Bilirubin Day 3                          &                 & double           & 297          \\
crea          & Serum creatinine Day 3                   &                 & double           & 3            \\
sod           & Serum sodium Day 3                       &                 & double           & 0            \\
ph            & Serum pH (arterial) Day 3                &                 & double           & 250          \\
glucose       & Glucose Day 3                            &                 & double           & 470          \\
bun           & BUN Day 3                                &                 & double           & 455          \\
urine         & Urine Output Day 3                       &                 & double           & 517          \\
adlp          & ADL Patient Day 3                        &                 & double           & 634          \\
adls          & ADL Surrogate Day 3                      &                 & double           & 310          \\
sfdm2         &                                          & 5               & integer          & 159          \\
adlsc         & Imputed ADL Calibrated to Surrogate      &                 & double           & 0           
\end{tabular}
\caption{A description of each variable in the SUPPORT dataset.}
\label{tab:support1}

\begin{table}[ht]
\centering
\begin{tabular}{cc}
\hline
\textbf{Variable} & \textbf{Levels}                      \\
\hline
sex      & female                                        \\
         & male                                          \\
dzgroup  & ARF/MOSF w/Sepsis                             \\
         & COPD                                          \\
         & CHF                                           \\
         & Cirrhosis                                     \\
         & Coma                                          \\
         & Colon Cancer                                  \\
         & Lung Cancer                                   \\
         & MOSF w/Malig                                  \\
dzclass  & ARF/MOSF                                      \\
         & COPD/CHF/Cirrhosis                            \\
         & Coma                                          \\
         & Cancer                                        \\
income   & under \$11k                                   \\
         & $11-$25k                                      \\
         & $25-$50k                                      \\
         & \textgreater{}\$50k                           \\
race     & white                                         \\
         & black                                         \\
         & asian                                         \\
         & other                                         \\
         & hispanic                                      \\
sfdm2    & no(M2 and SIP pres)                           \\
         & adl\textgreater{}=4 (\textgreater{}=5 if sur) \\
         & SIP\textgreater{}=30                          \\
         & Coma or Intub                                 \\
         & \textless{}2 mo. follow-up                   
\end{tabular}
\caption{A description of each level within each categorical variable in the dataset.}
\label{tab:support2}
\end{table}

As before, the first step is to visualize the incidence density. The
population-time plot for the SUPPORT data appears in Figure
\ref{fig:support-poptime}.
\textcolor{red}{This population-time plot is strange. The red dots never seem to go above 200.}

\begin{CodeInput}
R> data(support)
R> support_popTime <- popTime(support,
R+                            time = "d.time", event = "death")
R> plot(support_popTime)
\end{CodeInput}
\begin{figure}

{\centering \includegraphics{../figures/support-poptime-1}

\caption{\label{fig:support-poptime} Population-time plot for the SUPPORT data.}\label{fig:support-poptime}
\end{figure}
\end{CodeChunk}

Before we proceed with the analysis, we first remove hospital death as a
covariate, as this is directly informative of death and cannot be used
as a predictor. Then we select one observation at random. We will
estimate the cumulative incidence curve for this individual. The other
observations will be used to fit the hazard function.

\begin{CodeInput}
R> support <- dplyr::select(support, -hospdead)
R> 
R> # Randomly select one observation
R> indiv <- sample(1:nrow(support), size = 1)
\end{CodeInput}
\end{CodeChunk}

Our main objective is to estimate the cumulative incidence of death for
a given covariate profile. Given the number of covariates available, we
will do this using regularized estimation. As \pkg{casebase} makes use
of the \pkg{glmnet} package, we will use the function
\code{fitsmoothhazard.fit}, which follows the syntax from \code{glmnet}.
Specifically, we input a matrix \code{y}, which contains the time and
event variables, and a matrix \code{x}, which contains all other
variables we would like to include in the model. For the SUPPORT
dataset, all factor variables were converted into dummy variables.

We are now ready to estimate the hazard function. We use lasso
penalization; this is achieved by setting \code{alpha} to 1. Also, note
that the argument \code{formula_time} can be used to specify the
function form for time; in our example, we model time linearly.

\begin{CodeInput}
R> support_cb <- fitSmoothHazard.fit(support_train, y, 
R+                                   formula_time = ~ d.time, 
R+                                   family = "glmnet", 
R+                                   event = "death", time = "d.time", 
R+                                   alpha = 1, ratio = 100)
\end{CodeInput}
\end{CodeChunk}

Then, we take our model object and use the \code{absoluteRisk} function
to compute our estimate of the cumulative incidence function. Since
\code{glmnet} fitted the penalized logistic regression for a range of
penalty terms, we need to specify which one we want to use for our
prediction. We select \code{s = "lambda.1se"}, as recommended.

\begin{CodeInput}
R> time_points <- seq(1, max(support$d.time), by = 0.5)
R> support_risk <- absoluteRisk(support_cb,
R+                              time = time_points,
R+                              newdata = support_test,
R+                              s = "lambda.1se", method = "numerical")
\end{CodeInput}
\end{CodeChunk}

We will compare this curve to a regularized version of Cox regression,
and a Kaplan-Meier survival curve. Using the \pkg{glmnet} package, we
can fit the Cox regression model using regularized estimation.

\begin{CodeInput}
R> # Create survival object for glmnet
R> surv_glmnet <- survival::Surv(time = y[, "d.time"], 
R+                               event = y[, "death"])
R> 
R> support_cox <- glmnet::cv.glmnet(x = support_train, 
R+                                  y = surv_glmnet,
R+                                  family = "cox", alpha = 1)
\end{CodeInput}
\end{CodeChunk}

However, \pkg{glmnet} does not provide any functionality for estimating
the baseline hazard, and the functionalities in the \pkg{survival}
package do not work on the output from \code{cv.glmnet}. To circumvent
this issue and get an estimate of the baseline hazard, we followed the
following steps. First, we fitted a non-regularized Cox model using only
the variables selected by \pkg{glmnet}; the baseline hazard will be
estimated using this object. Then, we replaced the estimates of the
regression coefficients by the estimates from \pkg{glmnet}. Finally,
this modified \code{coxph} object is fed to the \code{survival::survfit}
function to obtain the estimated survival function.

\begin{CodeInput}
R> # Identifying selected covariates
R> nonzero_covariates_cox <- predict(support_cox, type = "nonzero", 
R+                                   s = "lambda.1se")
R> nonzero_coef_cox <- coef(support_cox, s = "lambda.1se")
R> # Creating a new dataset that only contains the covariates selected by glmnet
R> support_cox_data <- as.data.frame(cbind(y,
R+                                         support_train[, nonzero_covariates_cox$X1]))
R> # Fitting a cox model
R> support_cox <- survival::coxph(Surv(time = d.time, event = death) ~ ., 
R+                           data = support_cox_data)
R> # Replace the estimate with those obtained by glmnet
R> support_cox$coefficients <- nonzero_coef_cox@x
R> 
R> # Fitting CI curve for cox+glmnet
R> support_cox_risk <- survival::survfit(support_cox,
R+                                       newdata = as.data.frame(support_test),
R+                                       time = time_points,
R+                                       type = "breslow")
\end{CodeInput}
\end{CodeChunk}

We used the \pkg{survival} package to calculate the Kaplan-Meier
estimate of the survival function.

\begin{CodeInput}
R> km <- survival::Surv(time = support$d.time, 
R+                      event = support$death)
R> abKm <- survival::survfit(km ~ 1,
R+                           type = 'kaplan-meier')
\end{CodeInput}
\end{CodeChunk}

Now that our three cumulative incidence functions have been calculated,
we compare them all in Figure \ref{fig:abSC}. Both \pkg{glmnet} and
\pkg{casebase} have a lower estimated cumulative incidence by the end of
the study, when compared to Kaplan-Meier.
\textcolor{red}{It would be good to have an idea of why the case-base and Coxnet curves are so different. At the very least, I think we should compare the select covariates.}

{\centering \includegraphics{../figures/abSupportComparison-1}

\caption{\label{fig:abSC} Comparison of cumulative incidence curves between regularized casebase in black, regularized Cox regression in red and Kaplan-Meier in blue.}\label{fig:abSupportComparison}
\end{figure}
\end{CodeChunk}

\hypertarget{case-study-4stanford-heart-transplant-data}{%
\section{Case study 4--Stanford Heart Transplant
Data}\label{case-study-4stanford-heart-transplant-data}}

Although the previous case studies provide a broad overview of the
capabilities of \pkg{casebase}, they all have two properties in common:

\begin{itemize}
  \item All covariates are fixed, i.e. they do not change over time;
  \item The estimated hazard functions satisfy the "proportional hazard" assumption.
\end{itemize}

Case-base sampling has been used in the literature to study vaccination
safety, where the exposure period was defined as the week following
vaccination \citep{saarela2015case}. Hence, the main covariate of
interest, i.e.~exposure to the vaccine, was changing over time. In this
context, case-base sampling offers an efficient alternative to nested
case-control designs or self-matching.

The purpose of the next case study is to show how \pkg{casebase} can
also be used for survival analysis problems that do not share the
properties above. To this end, we will use the Stanford Heart Transplant
Data \citep[\citet{crowley1977covariance}]{clark1971cardiac}; we will
use the version available in the \pkg{surivival} package.

Recall the setting of this study: patients were admitted to the Stanford
program after meeting with their physician and determining that they
were unlikely to respond to other forms of treatment. After enrollment,
the program searches for a suitable donor for the patient, which can
take anywhere between a few days to almost a year. We are interested in
the effect of a heart transplant on survival; therefore, the patient is
considered exposed only after the transplant has occurred.

As before, we can look at the population-time plot for a graphical
summary of the event incidence. As we can see, most events occur early
during the follow-up period, and therefore we do not expect the hazard
to be constant.

\begin{center}\includegraphics{../figures/stanford-poptime-1} \end{center}

Since the exposure is time-dependent, we need to manually define the
exposure variable \emph{after} case-base sampling and \emph{before}
fitting the hazard function. For this reason, we will use the
\texttt{sampleCaseBase} function directly.

Next, we will compute the number of days from acceptance into the
program to transplant, and we use this variable to determine whether
each population-moment is exposed or not.

Finally, we can fit the hazard using various linear predictors.

Note that the fourth model (i.e. \texttt{fit4}) includes an interaction
term between exposure and follow-up time. In other words, this model no
longer exhibit proportional hazards. The evidence of non-proportionality
of hazards in the Stanford Heart Transplant data has been widely
discussed \citep{arjas1988graphical}.

We can then compare the goodness of fit of these four models using the
Akaike Information Criterion (AIC).

\begin{CodeOutput}
#> Model1 Model2 Model3 Model4 
#>    794    756    728    715
\end{CodeOutput}
\end{CodeChunk}

As we can, the best fit is the fourth model. By visualizing the hazard
functions for both exposed and unexposed individuals, we can more
clearly see how the hazards are no longer proportional.

\begin{center}\includegraphics{../figures/stanford-hazard-1} \end{center}

The non-proportionality seems to be more pronounced at the beginning of
follow-up than the end. Finally, we can turn these estimate of the
hazard function into estimates of the cumulative incidence functions.

\begin{center}\includegraphics{../figures/stanford-risk-1} \end{center}

Note that we can easily adapt the code above to the situation where a
patient receives a heart transplant at a point in time of interest, for
example after 30 days.

\begin{center}\includegraphics{../figures/stanford-1y-haz-1} \end{center}

We can then compare the 1-year mortality risk without transplant and
with transplant at 30 days.

\begin{CodeOutput}
#>         
#> 365 0.33
\end{CodeOutput}

\begin{CodeOutput}
#> [1] 0.22
\end{CodeOutput}
\end{CodeChunk}

As we can see, the risk estimate at 1-year is about 30\% lower if the
patient receives a heart transplant at 30 days.

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

In this article, we presented the \proglang{R} package \pkg{casebase}
that provides functions to fit smooth parametric hazards and estimate
cumulative incidence functions using case-base sampling. We outlined the
theoretical underpinnings of the approach, we provided details about our
implementation, and we illustrated the merits of the approach and the
package through four case studies.

As a methodological framework, case-base sampling is very flexible. This
flexibility has been explored before in the literature: for example,
Saarela and Hanley \citeyearpar{saarela2015case} used case-base sampling
to model a time-dependent exposure variable in a vaccine safety study.
As another example, Saarela and Arjas \citeyearpar{saarela2015non}
combined case-base sampling and a Bayesian non-parametric framework to
compute individualized risk assessments for chronic diseases. In the
case studies above, we explored this flexibility along two fronts. On
the one hand, we showed how splines could be used as part of the linear
predictor to model the effect of time on the hazard. This strategy
yielded estimates of the survival function that were qualitatively
similar to semiparametric estimates derived from Cox regression;
however, case-base sampling led to estimates that \emph{varying smoothly
in time}. On the other hand, we also displayed the flexibility of
case-base sampling by showing how it could be combined with penalized
logistic regression to perform variable selection. Similarly, our
package \pkg{casebase} provides the user with the functionality of
Generalized Additive Models. Moreover, case-base sampling can be
combined with quasi-likelihood estimation to fit survival models that
potentially account for the presence of over-dispersion. All of the
examples illustrate how the case-base sampling framework in general, and
the package \pkg{casebase} in particular, allows the user to fit a broad
family of survival functions.

As presented in Hanley \& Miettinen \citeyearpar{hanley2009fitting},
case-base sampling is comprised of three steps: 1) sampling a case
series and a base series from the study; 2) fit the log-hazard as a
linear function of predictors (including time); and 3) use the fitted
hazard to estimate the cumulative incidence curve. Accordingly, our
package provides functions for each step. Moreover, the simple interface
of the \code{fittingSmoothHazard} function resembles the \code{glm}
interface, which should look familiar to new users. Our modular approach
also provides a convenient way to extend our package for new sampling or
fitting strategies.

In the case studies above, we compared the performance of case-base
sampling with that of Cox regression and Fine-Gray models. In terms of
function interface, \pkg{casebase} uses a formula interface that is
closer to that of \code{glm}, in that the event variable is the only
variable appearing on the left-hand side of the formula. By contrast,
both \code{survival::coxph} and \code{timereg::comp.risk} use arrays as
the left-hand side that captures both the event type and time. Both
approaches to modeling yield user-friendly code. In terms of output,
both approaches differ significantly. Case-base sampling produces smooth
hazard and smooth cumulative incidence curves, whereas Cox regression
and Fine-Gray models produce step-wise cumulative incidence functions
and never explicitly model the hazard function. Qualitatively, we showed
that by using splines in the linear predictor, all three models yielded
similar curves. However, the smooth nature of the output of
\pkg{casebase} improves interpretation for consumers of these
predictions. In Table \ref{tab:compCBvsCox}, we provide a side-by-side
comparison between the Cox model and case-base sampling.

\begin{table}
\caption{\label{tab:compCBvsCox}Comparison between the Cox model and case-base sampling}
\centering
\begin{tabular}[t]{llp{5cm}}
\toprule
Feature & Cox model & Case-base sampling\\
\midrule
Model type & Semi-parametric & Fully parametric\\
Time & Left hand side of the formula & Right hand side (allows flexible modeling of time)\\
Cumulative incidence & Step function & Smooth-in-time curve\\
Non-proportional hazards & Interaction of covariates with time & Interaction of covariates with time\\
Model testing &  & Use GLM framework \newline (e.g.\ LRT, AIC, BIC)\\
\addlinespace
Competing risks & Difficult & Cause-specific CIFs\\
% Prediction & Kaplan-Meier-based & ROC, AUC, risk reclassification probabilities\\
\bottomrule
\end{tabular}
\end{table}

Our choice of modeling the log-hazard as a linear function of covariates
allows us to develop a simple computational scheme for estimation.
However, as a downside, it does not allow us to model location and scale
parameters separately like the package \pkg{flexsurv}. For example, if
we look at the Weibull distribution as parametrised in
\code{stats::pweibull}, the log-hazard function is given by
\[ \log h(t; \alpha, \beta) = \left[\log(\alpha/\beta) - (\alpha - 1)\log(\beta)\right] + (\alpha - 1)\log t,\]
where \(\alpha,\beta\) are shape and scale parameters, respectively.
Unlike \pkg{casebase}, the approach taken by \pkg{flexsurv} allows the
user to model the scale parameter as a function of covariates. Of
course, this added flexibility comes at the cost of interpretability: by
modeling the log-hazard directly, the parameter estimates from
\pkg{casebase} can be interpreted as estimates of log-hazard ratios. To
improve the flexibility of \pkg{casebase} at capturing the scale of a
parametric family, we could replace the logistic regression with its
quasi-likelihood counterpart and therefore model over- and
under-dispersion with respect to the logistic likelihood. We defer the
study of the properties and performance of such a model to a future
article.

Future work will look at some of the methodological extensions of
case-base sampling. First, to assess the quality of the model fit, we
may want to study the properties of the residuals (e.g.~Cox-Snell,
martingale). More work needs to be done to understand these residuals in
the context of the partial likelihood underlying case-base sampling. The
resulting diagnostic tools would then be integrated in this package.
Also, we are interested in extending case-base sampling to account for
interval censoring. This type of censoring is very common in
longitudinal studies, and many packages (e.g. \pkg{SmoothHazard},
\pkg{survival} and \pkg{Rstpm2}) provide functions to account for it.
Again, we hope to include any resulting methodology as part of this
package.

In future versions of the package, we also want to increase the
complement of diagnostic and inferential tools that are currently
available. For example, we would like to include the ability to compute
confidence intervals for the cumulative incidence curve. The delta
method or parametric bootstrap are two different strategies we can use
to construct approximate confidence intervals. Furthermore, we would
like to include functions to compute calibration and discrimination
statistics (e.g.~Brier score, AUC) for our models. Saarela and Arjas
\citeyearpar{saarela2015non} also describe how to obtain a posterior
distribution for the AUC from their model.

In conclusion, we presented the \proglang{R} package \pkg{casebase}
which implements case-base sampling for fitting parametric survival
models and for estimating smooth cumulative incidence functions using
the framework of generalized linear models. We strongly believe that its
flexibility and its foundation on the familiar logistic regression model
will make it appealing to new and established practioners.

\hypertarget{environment-details}{%
\section{Environment Details}\label{environment-details}}

This report was generated on 2020-01-29 23:33:21 using the following
computational environment and dependencies:

\begin{CodeOutput}
#> - Session info ---------------------------------------------------------------
#>  setting  value                       
#>  version  R version 3.6.2 (2019-12-12)
#>  os       Pop!_OS 19.10               
#>  system   x86_64, linux-gnu           
#>  ui       X11                         
#>  language en_US:en                    
#>  collate  en_US.UTF-8                 
#>  ctype    en_US.UTF-8                 
#>  tz       America/Toronto             
#>  date     2020-01-29                  
#> 
#> - Packages -------------------------------------------------------------------
#>  package     * version    date       lib
#>  assertthat    0.2.1      2019-03-21 [1]
#>  backports     1.1.5      2019-10-02 [1]
#>  broom         0.5.4      2020-01-27 [1]
#>  callr         3.4.1      2020-01-24 [1]
#>  casebase    * 0.2.1.9001 2020-01-27 [1]
#>  cellranger    1.1.0      2016-07-27 [1]
#>  cli           2.0.1      2020-01-08 [1]
#>  codetools     0.2-16     2018-12-24 [4]
#>  colorspace    1.4-1      2019-03-18 [1]
#>  cowplot     * 1.0.0      2019-07-11 [1]
#>  crayon        1.3.4      2017-09-16 [1]
#>  data.table    1.12.8     2019-12-09 [1]
#>  DBI           1.1.0      2019-12-15 [1]
#>  dbplyr        1.4.2      2019-06-17 [1]
#>  desc          1.2.0      2018-05-01 [1]
#>  devtools      2.2.1      2019-09-24 [1]
#>  digest        0.6.23     2019-11-23 [1]
#>  dplyr       * 0.8.3      2019-07-04 [1]
#>  ellipsis      0.3.0      2019-09-20 [1]
#>  evaluate      0.14       2019-05-28 [1]
#>  fansi         0.4.1      2020-01-08 [1]
#>  farver        2.0.3      2020-01-16 [1]
#>  forcats     * 0.4.0      2019-02-17 [1]
#>  foreach       1.4.7      2019-07-27 [1]
#>  fs            1.3.1      2019-05-06 [1]
#>  generics      0.0.2      2018-11-29 [1]
#>  ggplot2     * 3.2.1      2019-08-10 [1]
#>  glmnet      * 3.0-2      2019-12-11 [1]
#>  glue        * 1.3.1      2019-03-12 [1]
#>  gtable        0.3.0      2019-03-25 [1]
#>  haven         2.2.0      2019-11-08 [1]
#>  hms           0.5.3      2020-01-08 [1]
#>  htmltools     0.4.0      2019-10-04 [1]
#>  httr          1.4.1      2019-08-05 [1]
#>  iterators     1.0.12     2019-07-26 [1]
#>  jsonlite      1.6        2018-12-07 [1]
#>  kableExtra  * 1.1.0      2019-03-16 [1]
#>  knitr         1.27       2020-01-16 [1]
#>  labeling      0.3        2014-08-23 [1]
#>  lattice       0.20-38    2018-11-04 [4]
#>  lava          1.6.6      2019-08-01 [1]
#>  lazyeval      0.2.2      2019-03-15 [1]
#>  lifecycle     0.1.0      2019-08-01 [1]
#>  lubridate   * 1.7.4      2018-04-11 [1]
#>  magrittr    * 1.5        2014-11-22 [1]
#>  MASS          7.3-51.5   2019-12-20 [4]
#>  Matrix      * 1.2-18     2019-11-27 [4]
#>  memoise       1.1.0      2017-04-21 [1]
#>  mgcv          1.8-31     2019-11-09 [4]
#>  modelr        0.1.5      2019-08-08 [1]
#>  munsell       0.5.0      2018-06-12 [1]
#>  nlme          3.1-143    2019-12-10 [4]
#>  numDeriv      2016.8-1.1 2019-06-06 [1]
#>  pillar        1.4.3      2019-12-20 [1]
#>  pkgbuild      1.0.6      2019-10-09 [1]
#>  pkgconfig     2.0.3      2019-09-22 [1]
#>  pkgload       1.0.2      2018-10-29 [1]

>>>>

\usepackage{amsmath} \usepackage{longtable} \usepackage{graphicx} \usepackage{tabularx} \usepackage{float} \usepackage{booktabs} \usepackage{makecell} \usepackage{tabu} \DeclareUnicodeCharacter{2500}{-}

@@

\begin{center}\includegraphics{../figures/unnamed-chunk-1-1} \end{center}

@@

@@

\caption{\label{tab:unnamed-chunk-4}Mean }
\centering
\begin{tabular}[t]{>{\bfseries}lcc}
\toprule
Model & casebase & survreg\\
\midrule
Exponential & 0.81 (0.68, 0.96) & 0.81 (0.68, 0.96)\\
\cmidrule{1-3}
Gompertz & 0.78 (0.66, 0.93) & 0.80 (0.67, 0.95)\\
\cmidrule{1-3}
Weibull & 0.80 (0.67, 0.95) & 0.80 (0.65, 0.96)\\
\cmidrule{1-3}
Splines & 0.81 (0.68, 0.96) & --\\
\bottomrule
\multicolumn{3}{l}{\textit{Note:}}\\
\multicolumn{3}{l}{Median (Inter-quartile range) is given for Model Size.}\\
\end{tabular}
\end{table}

@@
